---
title: rowwise manipulation of data frame in R
author: ''
date: '2019-09-13'
slug: rowwise-manipulation-of-data-frame-in-r
categories: []
tags: [R]
---

The data frame is used for storing tabular data in R. It is the fundamental data structure in R, especially for tidyverse. Tidy data (data frame), in which each variable in a column and each observation in a row, is used wherever possible througout all the tidyverse packages. Working on column is a very natural and usual operation, and the key ideology in  [dplyr](https://dplyr.tidyverse.org/), a core package of tidyverse. But how to perform rowwise manipulation ? In this post, I will show how to do rowwise operation on data frames using both base R and tidyverse.

## sample data

```{r sample_data, message=FALSE}
library(tidyverse)
library(rvest)

url <- "https://www.pesmaster.com/arsenal/pes-2020/team/101/"

html <- read_html(url)

players <- html_nodes(html, ".squad-table") %>% 
  html_table() %>% 
  `[[`(1) %>%
  arrange(desc(Ovr)) %>% 
  slice(1:10) %>% select(Name, Pas:Dri) %>% 
  column_to_rownames("Name") %>% 
  t() %>% 
  as.data.frame()
names(players) <- stringi::stri_trans_general(names(players), "latin-ascii") %>% # convert to latin-ascii
  make.names() %>% # valid names
  str_replace_all(".*\\.+", "") # family name
```
We want to find the the most skillful player of each ability (Pas, Sht, Phy, Def, Spd, and Dri) of ten Arsenal players [in PES 2020](https://www.pesmaster.com/arsenal/pes-2020/team/101/).

## base R

### `apply()`

Once rowwise operation is mentioned, the first function comes mind is `apply()`.

```{r apply}
index <- apply(players, 1, which.max)
max_ability <- apply(players, 1, max)
skillful_player <- data.frame(
  name = names(players)[index],
  value = max_ability,
  ability = row.names(players)
)
```

We can use ggplot to visulaize the rusults

```{r apply-gg}
player_label <- data.frame(
  label = skillful_player$name,
  x = skillful_player$ability,
  y = 100
)
ggplot(skillful_player, aes(x = ability, y = value)) +
  geom_col(aes(fill = ability)) +
  geom_text(aes(x = x, y = y, label = label), data = player_label)
```

The result shows that the best dribbling player is the new player PEPE, and Ozil
is still the best passer, although he have not participated in anyt official 
match for a long time.

### `for` loop

Of course someone has to write loops

```{r loop}
index_loop <- vector(mode = "double", nrow(players))
for (i in seq_along(index_loop)) {
  index_loop[i] <- which.max(players[i, ])
}
max_ability_loop <- vector(mode = "double", nrow(players))
for (i in seq_along(max_ability_loop)) {
  max_ability_loop[i] <- max(players[i, ])
}
```

Apprantely, using for loop is more intuitive, but requires more typing.

### split, then apply and combine

The Next method is split the data frame by row then apply and combine it.

```{r split-apply}
players_split <- split(players, seq_len(nrow(players)))
max_ability <- sapply(players_split, max)
index <- sapply(players_split, which.max)
skillful_player_split <- data.frame(
  value = max_ability,
  name = names(players)[index],
  ability = row.names(players)
)

ggplot(skillful_player_split, aes(x = ability, y = value)) +
  geom_col(aes(fill = ability)) +
  geom_text(aes(x = x, y = y, label = label), data = player_label)
```

## tidyverse

### `pmap()` in purrr

`purrr::pmap()` iterate over multiple arguments simultaneously

```{r pmap}

which_max <- function(...) {
  which.max(c(...))
}
skillful_player_pmap <- players %>% 
  mutate(index = pmap_int(., which_max),
    value = pmap_int(., max),
    names = names(players)[index],
    ability = row.names(players)
  ) 


ggplot(skillful_player_pmap, aes(x = ability, y = value)) +
  geom_col(aes(fill = ability)) +
  geom_text(aes(x = x, y = y, label = label), data = player_label)
```

### `rowwise()` in dplyr

**dplyr** provides a function `rowwise()` to preform row-wise operations. However,
[as mentioned in this issue](https://github.com/tidyverse/dplyr/issues/3890#issuecomment-428320156),
we can not use tidyselect operation `:`, which means that all variables
must be explicitly provided for rowwise manipulation. `rowwise()` is not suitable 
while there are many varaibles.

```{r rowwise, eval=FALSE}
skillful_player_rowwise <- players %>% 
  rowwise() %>% 
  # mutate(value = max(Aubameyang:Kolasinac)) does not work well
  mutate(value = max(Aubameyang, Lacazette, Leno, Sokratis, 
    Luiz, Ozil, Torreira, Xhaka, Pepe, Kolasinac)
  )
```

Furthermore, we can also transpose the data frame first, and then use apply or **dplyr**
or `purrr::map()` to perform rowwise operation. Intuitively, this method is  more 
complicated than the method mentioned above, so here we not detail the code.

In summary, `purrr::pmap()` is using tidyverse and mmore easy to used as part of a pipe. 
The apply and for loop is more intuitive and efficiency (as shown in [here](https://speakerdeck.com/jennybc/row-oriented-workflows-in-r-with-the-tidyverse?slide=40), but requires more typing. `dplyr::rowwise()` is not suitable while there are many varaibles.

## Reference

- [Row-oriented workflows in R with the tidyverse](https://github.com/jennybc/row-oriented-workflows#readme)

