---
title: Rowwise operation in dplyr 1.0.0
author: ''
date: '2020-03-13'
slug: rowwise-operation-in-dplyr-1-0-0
categories: [R]
tags:
  - tidyverse
  - dplyr
---

I have written a post on [rowwise operation of data frame in 
R](https://www.choyang.me/post/rowwise-manipulation-of-data-frame-in-r/) a 
while ago. `purrr::pmap()` is recommended for rowwise operation in that post, since 
other methods have their own disadvantages. However, there will be a better rowwise operation support in **dplyr 1.0.0** ([will be released soon](https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-is-coming-soon/)), 
and it is very intuitive, simple, easy to use. 

## Basic

```{r rowwise, message=FALSE}
library(dplyr)
df <- tibble(x = 1:3, y = 2:4, z = 3:5)
df %>% rowwise() %>% mutate(m = mean(c(x, y, z)))
```

We can use tidy selection syntax to succinctly select any variables with
`c_across()`.

```{r across}
df %>% rowwise() %>% mutate(m = mean(c_across(everything())))

# equal to
df %>% rowwise() %>% mutate(m = mean(c_across(is.numeric)))
df %>% rowwise() %>% mutate(m = mean(c_across(x:z)))
```

`rowwise()` is behave somewhat similarly to the grouping variables passed to `group_by()`, we can preserve variables by `rowwise(<var_to_perserve>)`.

```{r rowwise-preserve}
# .before = x means new var `v` should locate before var `x`
df2 <- mutate(df, v = letters[1:3], .before = x) 
df2 %>% rowwise(v) %>% mutate(m = mean(x:z))
```

## row-wise summary funcitons in base R

For more efficient, we can use row-wise summary functions in base R 
`across()` is required for multiple 

```{r row-wise}
# use rowMeans 
df %>% mutate(m = rowMeans(across(everything())))

# equal to
mutate(df, m = rowMeans(df))
```

## Advanced usage

Run a function many times with different arguments

```{r simulation}
# example from dplyr vignette, rowwise
df <- tribble(
  ~ n, ~ min, ~ max,
    1,     0,     1,
    2,    10,   100,
    3,   100,  1000,
)

# list is required in here, since `mutate` has to return something in length 1
df %>% 
  rowwise() %>% 
  mutate(data = list(runif(n, min, max)))

# we can also use `purrr::pmap()`
df %>% mutate(data = purrr::pmap(., .f = runif))
```

More complicated problems, vary the function being called

```{r var-function}
# example from dplyr vignette, rowwise
df <- tribble(
   ~rng,     ~params,
   "runif",  list(n = 10), 
   "rnorm",  list(n = 20),
   "rpois",  list(n = 10, lambda = 5),
)

df %>% 
  rowwise() %>% 
  mutate(data = list(do.call(rng, params)))

# use purrr::map2, more complicated
df %>% mutate(data = purrr::map2(rng, params, ~ do.call(.x, .y)))
```

Fore more details see the vignette in dplyr.

## Reference 

- https://dplyr.tidyverse.org/dev/articles/rowwise.html





